---
title: "gaussianity"
output: 
date: "2023-04-07"
---

# NOTES

-   for MANOVA we need multivariate gaussianity for each group

-   for ANOVA we need univariate gaussianity for each group

-   we do not need multivariate gaussianity for the whole dataset since we do not want to perform tests on the mean for the whole dataset

# REFERENCES

-   <https://stats.stackexchange.com/questions/2492/is-normality-testing-essentially-useless>

# PRELIMINARY STUFF

```{r}
#directories
working_dir = "/Users/marcogalliani/Desktop/PISA-dev-local"
dataset_dir = "/Users/marcogalliani/Desktop/PISA-dev-local/data"
include_dir = paste(working_dir,"/local-src/include/Utilities.R",sep="")
#including utilities
source(include_dir)
#importing the dataset
pisa_data <- read.csv(file=paste(dataset_dir,"/pisa-woNA.csv",sep=""))

#select only ICT moments fot the moment
library(dplyr)
pisa_data <- pisa_data %>% dplyr::select(all_of(c("CNT",stu_ICT,sch_ICT)))
```

# CHECKING GAUSSIANITY

Reference: `Lab_4.R`

From `Lab_4.R` we have many options to test normality of data:

-   **Henze-Zikler's test**: initially it gives some weird errors about singularity, then after removing the variable RATCMP2 which had null estimated variance it begins to work

-   **Shapiro-Wilk test family**: It is not possible to perform these kind of test on our data, since they are limited to 5000 observations

-   We can reject Gaussianity also with univariate tests like the ANDERSON-DARLING test. However, we can only reject gaussianity and not accept it, since gaussianity of the conditions is a necessary condition for the multivariate gaussianity.

## Problems

-\> Null estimated variance of RATCMP2

```{r}
image(cov(select_if(pisa_data,is.numeric)))
pisa_data$RATCMP2 <- NULL #RATCMP2 seems has null estimated variance and gives problems
image(cov(select_if(pisa_data,is.numeric)))
```

-\> Unbalanced number of observations for Spain (too many)

```{r}
#number of observations b y group
pisa_data %>% count(CNT,sort=TRUE)
#here we can see that we have many observations about Spain. Maybe we can try excluding it to have a more balanced data
pisa_data <- pisa_data %>% filter(CNT!="ESP")
```

## Testing

In `result` we can see also the results of the univariate tests on gaussianity

```{r}
library(MVN)
#multivariate Shapiro-Wilk test: not possible to perform
#sample size is too big -> maybe perform the test sampling data from the datset
result <- mvn(data = pisa_data, subset = "CNT", mvnTest = "hz",univariateTest = "AD")
#to explore the options of mvn
#help(mvn)
```

# REMOVING OUTLIERS

```{r}
M <- colMeans(select_if(pisa_data,is.numeric))
S <- cov(select_if(pisa_data,is.numeric))
#computing mahlanobis distance
d2 <- matrix(mahalanobis(select_if(pisa_data,is.numeric), M, S))
#removing over a certain distance from the mean
pisa_wo_outliers<- pisa_data[which(d2 <= 10), ] #too many observations removed with 8

#testing again on the dataset wo outliers
result_wo_ouliers <- mvn(data = pisa_wo_outliers, subset = "CNT", mvnTest = "hz",univariateTest = "AD")
```

# TRANSFORMING

```{r}
library(car)
bcn_param <- powerTransform(select_if(pisa_wo_outliers,is.numeric), family = "bcnPower") #take long
trasformed_data <- bcnPower(U=select_if(pisa_wo_outliers,is.numeric),lambda=bcn_param$lambda, gamma=bcn_param$gamma)
#add the CNT column
transformed_data["CNT"] <- pisa_wo_outliers$CNT

#testing on transformed data
result_wo_ouliers <- mvn(data = transformed_data, subset = "CNT", mvnTest = "hz",univariateTest = "AD")
```
