---
title: "R Notebook"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---
# TO DO
- transforming the dataset (if necessary)
- Ridge and Lasso (?)
- Plots (if number of variables permits it)
- introduction of groups (for LMM)

```{r, setup}
#DIRECTORIES
rm(list=ls())
graphics.off()
root_proj_dir = "../../"
dataset_path = paste(root_proj_dir,"data/pisa_wPV_grouped_bysch.csv",sep="")
include_path = paste(root_proj_dir,"src/include/Utilities.R",sep="")
#INCLUDING UTILITIES
source(include_path)
#IMPORTING THE DATASET
data <- read.csv(file=dataset_path)
head(data)
library(mvtnorm)
library(MASS)
library(car)
library(rgl)
library(leaps)
library(ISLR)
library(glmnet)


```

```{r}
data$X <- NULL
data$schID <- NULL
data$CNT <- as.factor(data$CNT)
data$CNTSCHID <- as.factor(data$CNTSCHID)
data$CNTSTUID <- as.factor(data$CNTSTUID)
data <- data[,23:74] #excluding target variables
head(data)
```

# choice of the model
```{r}
total_model = c("ATTLNACT","EMOSUPS","COMPETE","GFOFAIL","EUDMO","RESILIENCE","BELONG","BEINGBULLIED",
                "PERFEED","CREACTIV","STRATIO","SCHSIZE","CLSIZE","EDUSHORT","STAFFSHORT","STUBEHA","TMINS",
                "JOYREAD","CULTPOSS","HEDRES","SCREADCOMP","COMPICT","ICTCLASS","ICTHOME","ICTOUTSIDE","ICTRES",
                "ENTUSE","HOMESCH","USESCH","INTICT","AUTICT")
response_variable_total = "SWBP"

########
tech <- c("ICTCLASS","ICTHOME","ICTOUTSIDE","ICTRES","ENTUSE","HOMESCH","USESCH","INTICT","AUTICT")
tech2 <- c("AUTICT","INTICT")
response_variable_tech = "COMPICT"
########

culture = c("JOYREAD","CULTPOSS","HEDRES","SCREADCOMP")
TMINS=data$MMINS+data$LMINS
response_variable_culture = "TMINS"
########

psychology = c("ATTLNACT","EMOSUPS","COMPETE","EUDMO","GFOFAIL","RESILIENCE","BELONG","BEINGBULLIED","PERFEED")
response_variable_psy = "SWBP"
########

school_prof = c("STRATIO","SCHSIZE","CLSIZE","EDUSHORT","STAFFSHORT","STUBEHA")
response_variable_school = "CREACTIV"
########

school_prof_2 = c("SCHSIZE","CLSIZE","STAFFSHORT")
response_variable_school_2 = "STRATIO"
########################


vars = psychology
response_variable = response_variable_psy
number_of_covariates = length(vars)
number_of_covariates
```

# run the model
```{r}
formula <- paste(paste(response_variable,"~"), paste(vars, collapse = "+"))
formula
linear_model <- lm(formula,data)
summary(linear_model)
vif(linear_model)
```

#  Verify assumptions (used for inference and estimate of parameters)
```{r, warning=FALSE, message=FALSE}
# * Gaussianity
# * Homoschedasticity
plot(linear_model)

# Comment
#   1. We want to see no pattern: a cloud around the zero
#   2. We want to see a good fit on the line
#   3. Again, we want to see no pattern
#   4. We have the iso-lines of the Cook distance: we can identify the outliers

shapiro.test(residuals(linear_model))

```


# Variable selection

## **1.** Best Subset Selection (exhaustive search) 
```{r, warning=FALSE, message=FALSE}
# Best Subset Selection
formula_reg = formula(paste(response_variable,"~."))
regfit.full = regsubsets(formula_reg, data=data,really.big = T)
summary(regfit.full)

# Best Subset Selection: we say when we stop
nv_max = number_of_covariates
regfit.full = regsubsets(formula_reg, data=data, nvmax=nv_max,really.big = T)
summary(regfit.full)

reg.summary = summary(regfit.full)

# Which one we choose:
reg.summary$which

# R-squared
reg.summary$rsq

# R.adj^2
reg.summary$adjr2

# SSres (residual sum of squares)
reg.summary$rss

# Plots
par(mfrow=c(1,3))
plot(reg.summary$rsq, xlab="Number of Variables", ylab="R-squared", type="b")
plot(reg.summary$adjr2, xlab="Number of Variables", ylab="Adjusted RSq", type="b")
plot(reg.summary$rss, xlab="Number of Variables", ylab="RSS", type="b")

# We want the model with max r.adj^2 so we extract the coefficients of that model
# Note: ind = how many coefficients has the model
ind = which.max(reg.summary$adjr2)
coef(regfit.full, ind)

# Graphical table of best results
par(mfrow=c(1,2))
plot(regfit.full, scale="r2", main="Exhaustive search")
plot(regfit.full, scale="adjr2", main="Exhaustive search")

```

## **2.** Forward and Backward Stepwise Selection
```{r, warning=FALSE, message=FALSE}
# Forward
regfit.fwd = regsubsets(Y~.,data=data,nvmax=nv_max,method="forward")
summary(regfit.fwd)

# Plot
par(mfrow=c(1,3))
plot(summary(regfit.fwd)$rsq, xlab="Number of Variables", ylab="R-squared", type="b")
plot(summary(regfit.fwd)$adjr2, xlab="Number of Variables", ylab="Adjusted RSq", type="b")
plot(summary(regfit.fwd)$rss, xlab="Number of Variables", ylab="RSS", type="b")

par(mfrow=c(1,2))
plot(regfit.fwd,scale="r2",main="Forward Stepwise Selection")
plot(regfit.fwd,scale="adjr2",main="Forward Stepwise Selection")

# Backward
regfit.bwd = regsubsets(Y~.,data=data,nvmax=nv_max,method="backward")
summary(regfit.bwd)

# Plot
par(mfrow=c(1,3))
plot(summary(regfit.bwd)$rsq, xlab="Number of Variables", ylab="R-squared", type="b")
plot(summary(regfit.bwd)$adjr2, xlab="Number of Variables", ylab="Adjusted RSq", type="b")
plot(summary(regfit.bwd)$rss, xlab="Number of Variables", ylab="RSS", type="b")

par(mfrow=c(1,2))
plot(regfit.bwd,scale="r2",main="Backward Stepwise Selection")
plot(regfit.bwd,scale="adjr2",main="Backward Stepwise Selection")

```

## **3.** Comparison
```{r, warning=FALSE, message=FALSE}
coef(regfit.full,7) # Exhaustive search
coef(regfit.fwd,7) # Forward Stepwise Selection
coef(regfit.bwd,7) # Backward Stepwise Selection
```

## **4.** K-fold-cross-validation (exhaustive search)
```{r, warning=FALSE, message=FALSE}
k = 10
folds = sample(1:k,nrow(data),replace=TRUE)
table(folds)

# Function that performs the prediction for regsubsets
predict.regsubsets = function(object,newdata,id){
  form  = as.formula(object$call[[2]])
  mat   = model.matrix(form,newdata)
  coefi = coef(object,id=id)
  xvars = names(coefi)
  mat[,xvars]%*%coefi
}
p = number_of_covariates
cv.errors = matrix(NA,k,p, dimnames=list(NULL, paste(1:number_of_covariates)))

for(j in 1:k){
  best.fit = regsubsets(formula_reg,data=data[folds!=j,],nvmax=number_of_covariates,really.big=T)
  for(i in 1:p){
    pred = predict(best.fit,data[folds==j,],id=i)
    cv.errors[j,i] = mean( (data$Y[folds==j]-pred)^2 )
  }
}

cv.errors

root.mean.cv.errors = sqrt(apply(cv.errors,2,mean)) # average over the columns
root.mean.cv.errors

# Plot
plot(root.mean.cv.errors,type='b')
points(which.min(root.mean.cv.errors),
       root.mean.cv.errors[which.min(root.mean.cv.errors)], col='red',pch=19)
which.min(root.mean.cv.errors)

# Estimation on the full dataset
reg.best = regsubsets(Y~.,data=data, nvmax=number_of_covariates)
coef(reg.best,10)

```
# Inference on Betas
```{r, warning=FALSE, message=FALSE}
## Assuption: Eps ~ N(0, sigma^2)
## Test (Fisher):
##    H0: (beta1, beta2) == (0, 0)
##    H1: (beta1, beta2) != (0, 0)
r = linear_model$rank - 1  # number of regressors

# needs to be specified
kk=9
linearHypothesis(linear_model, cbind(rep(0,kk),diag(kk)), rep(0,kk))
linear_model
# Comment
#   Pr(>F) = final p-value in summary(fm)
```

#  Bonferroni Intervals 
```{r, warning=FALSE, message=FALSE}
alpha = 0.05
qT = qt(1-alpha/(2*p), n-(r+1))

vals = c(2,3)
C = diag(3)
Bf = c()
for(j in vals){
  Bf = rbind(Bf,
         c( (C %*%coefficients(linear_model))[j]-sqrt((C %*%vcov(linear_model) %*% t(C))[j,j])*qT,
            (C %*%coefficients(linear_model))[j]+sqrt((C %*%vcov(linear_model) %*% t(C))[j,j])*qT))
}

Bf

# Generic beta_j (p, r, n, alpha generici)
# beta_j = c(coefficients(fm)[j]-sqrt(vcov(fm)[j,j])*qt(1-alpha/(2*p), n-(r+1)),
#            coefficients(fm)[j]+sqrt(vcov(fm)[j,j])*qt(1-alpha/(2*p), n-(r+1)))

# Alternatively: Bonferroni's correction
confint(fm, level= 1-alpha/p)[vals,]
```




# variable transformation-tech
```{r}
attach(data)
AUT2=AUTICT^2
INT2=INTICT^2
AUT3=AUTICT^3
INT3=INTICT^3
tec3=lm(COMPICT ~ AUTICT+INTICT+INT2+INT3+AUT3)
summary(tec3)
#R_adj^2 0.5839 si sta alzando ma di poco -> ordine salirebbe troppo prima di raggiungere R^2 accettabile -> overfitting 
vif(tec3) #vif ancora buono
detach(data)
```

###
Questi sono solo possibili esempi -> anche cambiando variabili non si nota un miglioramento dell'R_adj^2

=>variabili non si spiegano l'una con l'altra
nonostante pvalue basso modello non molto buono
Idem per ciò che segue





# Provare con gli alberi
```{r}

attach(data)


tree.tec <- tree(COMPICT ~ ICTCLASS+ICTHOME+ICTOUTSIDE+ICTRES+ENTUSE+AUTICT,data)
summary(tree.tec)

plot(tree.tec)
text(tree.tec,pretty=0)

cv.tec <- cv.tree(tree.tec)
plot(cv.tec$size,cv.tec$dev,type='b',xlab='size',ylab='deviance')

#pruning per ridurre albero anche se solo una variabile considerata
prune.tec <- prune.tree(tree.tec,best=4)
plot(prune.tec)
text(prune.tec,pretty=0)

#considera solo autict 

detach(data)

```

