---
title: "R Notebook"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

## Settings

```{r}
#loaded librarires
library(dplyr)
library(psych) #for KMO test and principal()
library(car) #to apply transformations
library(MVN) #to perform multivariate gaussianity check
library(GGally) #for ggcorr
library(ggplot2)
library(elasticnet) #trying out sparse principal component analysis [spca()]
library(tidyverse)
```

```{r, setup}
#DIRECTORIES
root_proj_dir = "../../../"
dataset_dir = paste(root_proj_dir,"/data/pisa_data_final.csv",sep="")
include_dir = paste(root_proj_dir,"/src/include/Utilities.R",sep="")
#INCLUDING UTILITIES
source(include_dir)
#IMPORTING THE DATASET
pisa_data <- read.csv(file=dataset_dir)
colnames(pisa_data)
head(pisa_data)
```
# Modify dataset
```{r}
pisa_data$X <- NULL
pisa_data$CREACTIV <- as.factor(round(pisa_data$CREACTIV))
pisa_data$CNT <- as.factor(pisa_data$CNT)
pisa_data$SCHLTYPE <- as.factor(pisa_data$SCHLTYPE)
head(pisa_data)
# remove PV
pisa_data <- pisa_data %>% select(-starts_with("PV"))
# LMINS and MMINS averaged
pisa_data$SMINS = (pisa_data$LMINS+pisa_data$MMINS)/2

head(pisa_data)
```
#standardize the variables

```{r}
transformed_data <- as.data.frame(scale(select_if(pisa_data,is.numeric)))
data <- transformed_data

col_names_pisa = colnames(pisa_data)
col_names_data = colnames(data)
facotor_vars <- setdiff(col_names_pisa,col_names_data)
facotor_vars # those needs to be added later
```
# Remove variables that are already ok (numeric) + psyc that will be dealt differently
```{r}
remove_vars = c("EUDMO","SWBP","GFOFAIL", "RESILIENCE", "COMPETE",  #Psychological Well-being
                  "PERCOOP","PERCOMP","EMOSUPS","BELONG","BEINGBULLIED", #Social Well-Being
                  "WEALTH","HOMEPOS","BFMJ2","BMMJ1","HISCED","HISEI", # Family: ESCS is enpough
                  "LMINS","MMINS" #just the average
                  
)

original_vars = c("ESCS")
                  
```

# Groups
```{r}
grouped_variables <-list()
#list of grouped variables

grouped_variables[["ICT at home"]] <- c("ICTHOME","ICTRES","ENTUSE","HOMESCH")
grouped_variables[["Relationship with ICT"]] <- c("AUTICT","COMPICT","INTICT","ENTUSE", "HOMESCH", "USESCH")
grouped_variables[["ICT at school"]] <- c("ICTCLASS","ICTOUTSIDE","USESCH", "ICTSCH","RATCMP1")

##Culture
grouped_variables[["Cultural possesion"]] <- c("CULTPOSS","HEDRES")
grouped_variables[["Attitude towards school"]] <- c("SMINS","STUBEHA","ATTLNACT") #not good pca
grouped_variables[["Reading"]] <- c("JOYREAD", "SCREADCOMP")

##Teaching
grouped_variables[["Teachers' degree"]] <- c("PROAT5AB","PROAT5AM","PROAT6")
grouped_variables[["Teacher support"]] <- c("TEACHINT","TEACHSUP","STIMREAD","TEACHBEHA","PERFEED") #not good pca

##School
grouped_variables[["School size"]] <- c("STRATIO","SCHSIZE","CLSIZE") #not good pca
grouped_variables[["Equipment of the school"]] <- c("EDUSHORT","STAFFSHORT")

#groups
group_list <- names(grouped_variables)

names_grouped =c( (unlist(grouped_variables)),original_vars,facotor_vars,remove_vars)
col_names_data = colnames(pisa_data)
difference1 <- setdiff(col_names_data,names_grouped)
difference1
```

# Correlation within groups
Some graphs to validate the choice made while grouping the variables

```{r}
p_ <- GGally::print_if_interactive

#overall correlation overview
ggcorr(select_if(pisa_data, is.numeric),
       label = TRUE,
       label_size = 0.5)

for(group_name in group_list){
  
  p_(ggcorr(data[,grouped_variables[[group_name]]],
         label=TRUE,label_size = 2) +
    ggtitle(group_name))
}

```


## Selecting the number of factors

the following methods are useful to select the initial number of factors. The final choice should be made based on the portion of variance explained

(Note that scree plots are contained in parallel analysis)

```{r}
nfactors <- list()

for(group in group_list){
  #scree plot
  #scree(pisa_data[,grouped_variables[[group]]], 
  #      pc=TRUE, 
  #      main = paste("Scree plot:", group))
  
  #parallel analysis
  parallel.an <- fa.parallel(data[,grouped_variables[[group]]], 
              main = paste("Parallel Analysis Scree Plots:",group))
  
  nfactors[[group]] <- parallel.an$ncomp
  
  #a method from psych package with many more criteria
  #nfactors(cor(pisa_data[,grouped_variables[[group]]]),
  #         title = paste("Number of factors", group),
  #         plot=FALSE)
}
```




## Fitting the models

-   principal(): performs PCA in a more sophisticated way, using rotations to inrease the interpretability of the components. See the documentation in this folder [package psych]

```{r}
# rotated PCA
fit_PCA_rotated <- list()

#fitting the models
for(group in group_list){
  
  #advanced PCA
  fit_PCA_rotated[[group]] <- principal(r = cor(data[,grouped_variables[[group]]]),
                               nfactors = nfactors[[group]], 
                               rotate = "promax")
}
```


-   variance explained (it is the average of the communalities)

```{r}
cum_var = list()
for(group in group_list){
  cat("\n-----------Group:", group, "-----------\n")
  
  #number of factors and components
  cat("\nNumbr of components: ", nfactors[[group]])
  cat("-> fit: ", fit_PCA_rotated[[group]]$fit,"\n")
  cat("-> summary:\n")
  cum_var[[group]]=sum(fit_PCA_rotated[[group]]$Vaccounted["Proportion Var",])
  print(cum_var[[group]])
  #print(fit_PCA_rotated[[group]]$Vaccounted)

}
```
# Adjust number of components
```{r}

```



# Visualize the loadings:
```{r}

group_list
#Principal Component method
for(group in group_list){
  for(i in 1:nfactors[[group]]){
    barplot(fit_PCA_rotated[[group]]$loadings[,i], main = paste("factor",group,i),las=2,cex.names=0.7)
  }
}
```

# compute scores dataset
```{r}
scores <- function(fit,data){
  return(as.data.frame(as.matrix(data)%*%fit$loadings))
}

colnames(pisa_data)

scores_data = list()
for(group in group_list){
    scores_data[group] = list(scores(fit_PCA_rotated[[group]],pisa_data[,grouped_variables[[group]]]))
}

scores_data=data.frame(scores_data)
```











