---
title: "PISA dataset na exploration"
output: 
editor_options: 
  chunk_output_type: inline
---

# POOLS OF VARIABLES OF INTEREST

Here I select some variables about potentially interesting themes

Some details about the computation of the indexes

docs: <https://www.oecd-ilibrary.org/sites/0a428b07-en/index.html?itemId=/content/component/0a428b07-en>

```{r}
#Anagraphics

#Technology
stu_ICT = c("ICTHOME", "ICTSCH", "ICTRES", "ENTUSE", "HOMESCH","USESCH","INTICT","COMPICT","AUTICT","SOIAICT","ICTCLASS","ICTOUTSIDE")
sch_ICT = c("RATCMP1", "RATCMP2")

#Learning time
stu_LT = c("MMINS","LMINS","SMINS","TMINS")

#Economic status
stu_WEALTH = c("ESCS","WEALTH","HOMEPOS")

#Culture
stu_CULT = c("CULTPOSS","HEDRES")

#Well-being
stu_WB = c("TEACHSUP","EMOSUPS","COMPETE","GFOFAIL","EUDMO","SWBP","RESILIENCE","BELONG","BEINGBULLIED","PERCOMP","PERCOOP","BODYIMA")

#Attitude to global issues
stu_GLOBAL = c("GCSELFEFF","GCAWARE","ATTIMM","INTCULT","COGFLEX","RESPECT","AWACOM","GLOBMIND","DISCRIM")

#Concern for the future
stu_FUT = c("MASTGOAL","WORKMAST","INFOCAR","INFOJOB1","INFOJOB2")

#Family
stu_FAM = c("CURSUPP","EMOSUPP","PQSCHOOL","PASCHPOL","PRESUPP")
```

```{r}
#complete vectors
stu_vars = c(stu_ICT,stu_WB,stu_GLOBAL,stu_LT,stu_FUT,stu_FAM,stu_CULT,stu_WEALTH)
sch_vars = c(sch_ICT)
teach_vars = c()
```

# PREPROCESSING

```{r}
working_dir = "/Users/marcogalliani/Desktop/PISA-dev-local"
dataset_dir = "/Users/marcogalliani/Desktop/PISA-dev-local/data/PISA-dataset"
setwd(working_dir)
```

Data loading

```{r}
library(intsvy) # package to analyze PISA dataset

Europe = c("AUT","BEL","BGR","HRV","QCY","CZE","DNK","EST","FIN","FRA","DEU","GRC","HUN","IRL","ITA","LVA","LTU","LUX","NLD","POL","PRT","ROU","SVK","SVN","ESP","SWE","GBR")
#selecting and merging
pisa_europe <- pisa.select.merge(folder= dataset_dir,
                               school.file="CY07_MSU_SCH_QQQ.sav", 
                               student.file="CY07_MSU_STU_QQQ.sav",
                               student= stu_vars,
                               school = sch_vars,
                               countries = Europe)    
#pisa_world <- pisa.select.merge(folder= dataset_dir,school.file="CY07_MSU_SCH_QQQ.sav",  student.file="CY07_MSU_STU_QQQ.sav",student= stu_vars,school = sch_vars)   

```

Removing replicate weights (for simplicity: replicate weights have no NAs)

I've also decided to remove PVs since they have less NAs and I want to focus on other variables. The analysis on PVs will be made later.

Removing IDs, W_FSTUWT and W_FSTUWT_SC_SUM for the moment

```{r}
library(dplyr)
#pisa_world <- pisa_world %>% select(-starts_with("W_FSTU")) 
pisa_europe <- pisa_europe %>% select(-starts_with("W_FSTU")) 

#pisa_world <- pisa_world %>% select(-starts_with("PV")) 
pisa_europe <- pisa_europe %>% select(-starts_with("PV"))

#pisa_world <- pisa_world %>% select(-starts_with("PV")) 
pisa_europe <- pisa_europe %>% select(-starts_with("PV"))

#pisa_world <- pisa_world %>% select(-c("CNTRYID","CNTSCHID","CNTSTUID","BOOKID")) 
pisa_europe <- pisa_europe %>% select(-c("CNTRYID","CNTSCHID","CNTSTUID","BOOKID")) 
```

# GENERAL INSPECTION

```{r}
#select the variables of interest
inspect_europe <- pisa_europe[c("CNT",stu_ICT,sch_ICT,stu_CULT,stu_FAM,stu_WB,stu_WEALTH)]
#inspect_world <- pisa_world[c("CNT",stu_ICT,sch_ICT,stu_CULT,stu_FAM,stu_WB,stu_WEALTH)]
```

Inspecting the presence of NAs

docs: <https://cran.r-project.org/web/packages/naniar/vignettes/naniar-visualisation.html>

```{r}
# install.packages("naniar")
library(naniar)
# plot the NAs by columns: we see that there are patterns for NAs (i.e. same observations have NAs in different features)
vis_miss(inspect_europe,warn_large_data = FALSE)
#vis_miss(inspect_world,warn_large_data = FALSE)
```

```{r}
library(ggplot2)
gg_miss_var(inspect_europe,show_pct = TRUE) 
#gg_miss_var(inspect_world,show_pct = TRUE) 
```

See the combinantions of variables with more NAs

```{r}
library(UpSetR)
# help(gg_miss_upset)
gg_miss_upset(inspect_europe,nsets = 10)
#gg_miss_upset(inspect_world,nsets = 10)
# this graph tells us which combination of variables has the most common NA
```

Counting NAs by countries to see if we have countries with more NAs that we can exclude

```{r}
# help(gg_miss_case)
#visually
#gg_miss_case(inspect_world,facet = CNT)

#analytically
#na_count_by_group <- aggregate(. ~ inspect_world$CNT,inspect_world,function(x) { sum(is.na(x)) },na.action = NULL) 

##add nations count to the dataframe
#nation_count <- inspect_world %>% group_by(CNT) %>% summarise(total_count = n())
#na_count_by_group["nation count"] <- nation_count$total_count
```

```{r}
#gg_miss_fct(x = inspect_world,fct = CNT)
gg_miss_fct(x = inspect_europe,fct = CNT)
```

# SELECTION OF COUNTRY AND VARIABLES

```{r}
inspect_df_NA <- function(df){
  plot1 <- vis_miss(df,warn_large_data = FALSE)    
  plot2 <- gg_miss_upset(df,nsets = 10)
  plot3 <- gg_miss_fct(x = df,fct = CNT)
  return(list(plot1,plot2,plot3))
}
```

Select variables

```{r}
# BY PERCENTAGES
# na_europe <- round(colSums(is.na(inspect_europe))/dim(inspect_europe)[1],digits = 1)
# na_world <- round(colSums(is.na(inspect_world))/dim(inspect_world)[1],digits = 1)

# threshold = 0.25
# europe_sel <- inspect_europe[na_europe < threshold]
# world_sel <- inspect_world[na_world < threshold]

# BY SPECIFIC VARIABLES
europe_sel <- inspect_europe %>% select(-c("SOIAICT","PRESUPP","CURSUPP","EMOSUPP","PASCHPOL","PQSCHOOL","BODYIMA","SWBP")) 
#see the update
plots <- inspect_df_NA(europe_sel)
plots
```

Select country

```{r,figures.side,fig.show='hold'}
europe_sel <- europe_sel %>% filter(!(CNT %in% c("AUT","BEL","BGR","DEU","GBR","NLD","PRT","ROU","DEU")))
#see the update
plots <- inspect_df_NA(europe_sel)
plots
```

# TREATMENT OF NAs

```{r}
# 1) first option removing NAs with na.omit()
# europe_woNA <- na.omit(europe_sel)
# 2) filling with mean or median
library(tidyr)
europe_woNA <- europe_sel %>% group_by(CNT) %>% mutate_if(is.numeric, ~replace_na(.,mean(., na.rm = TRUE)))
#check if we have removed NAs
vis_miss(europe_woNA,warn_large_data = FALSE)
```

# SAVING DATA

```{r}
saving_dir = paste(working_dir,"/data/pisa-woNA.csv",sep="")
write.csv(europe_woNA,file=saving_dir)
```

# OBSERVATIONS AND TODO

-   analysis for plausible values
-   try filling with different methods
