---
title: 'One-way MANOVA, dataset grouped by school'
output: 
editor_options: 
  chunk_output_type: inline
---

# PRELIMINARY STUFF

```{r}
library(dplyr)
library(MVN)
library(car)
library(heplots)
library(mvnormtest)
library(gplots) #to plot text in graphics plot
```

```{r, setup}
#DIRECTORIES
root_proj_dir = "../../"
dataset_dir = paste(root_proj_dir,"/data/pisa-woNA_school_final.csv",sep="")
include_dir = paste(root_proj_dir,"/src/include/Utilities.R",sep="")
#INCLUDING UTILITIES
source(include_dir)
#IMPORTING THE DATASET
pisa_data <- read.csv(file=dataset_dir)
```

```{r}
#some adjustments on the data
pisa_data$X <- NULL
pisa_data$schID <- NULL
pisa_data$CNT <- as.factor(pisa_data$CNT)
```

# ONE-WAY MANOVA

```{r}
included_variables <-list()
#list of grouped variables
included_variables[["tec"]] <- c("ICTCLASS","ICTHOME","ICTOUTSIDE","ICTRES","AUTICT","COMPICT","INTICT","ENTUSE","HOMESCH","USESCH")
included_variables[["psi"]] <- c("ATTLNACT","EMOSUPS","COMPETE","EUDMO","GFOFAIL","SWBP","RESILIENCE","BELONG","BEINGBULLIED","PERFEED")
included_variables[["clt"]] <- c("JOYREAD","CULTPOSS","HEDRES","SCREADCOMP","LMINS","MMINS")
included_variables[["fam"]] <- c("WEALTH","ESCS","HOMEPOS","BFMJ2","BMMJ1","HISCED","HISEI")
included_variables[["tch"]] <- c("TEACHINT","TEACHSUP","STIMREAD")
included_variables[["sch"]] <- c("PERCOMP","PERCOOP","ICTSCH","RATCMP1")
```

Fitting the MANOVA model

```{r}
fit <- list()
fit_summaries <- list()
for(group in names(included_variables)){
  #recall that included_variables is a list of group of variables
  #here I'm accessing to one group at a time
  formula_str <- paste(included_variables[[group]], collapse=",")
  formula_str <- paste("cbind(",formula_str,")",sep = "")
  formula_str <- paste(formula_str,"CNT",sep = "~")
  #fitting the manova model
  fit[[group]] = manova(formula = as.formula(formula_str), data = pisa_data)
}
```

All the fits are stored in the \`fit\` object, they can be accessed with the following code (I'm looking for options to visualize them all together). However all variables seems to be influential

```{r}
#pick up one group from "tec" "psi" "clt" "fam" "tch" "sch"
group = "sch"
#result of the MANOVA test
print("result of the MANOVA test")
summary(fit[[group]])
#checking what country influenced the most the result of the test
print("checking what country influenced the most the result of the test")
summary.aov(fit[[group]])
```

## Checking the assumptions

(1) multivariate gaussianity in the groups: discussed extensively in \`src/gaussianity/gaussianity_by_schools.Rmd\`, here I just report a test

```{r}
normality_by_cntry <- list()
for(group in names(included_variables)){
  normality_by_cntry[[group]] <- mvn(data = subset(pisa_data,select = c(included_variables[[group]],"CNT")), subset = "CNT", mvnTest = "hz",univariateTest = "SW")
}
```

2.  same covariance structure (homoschedasticity)

```{r}
#TODO: not so important
```

# BONFERRONI CI

We want to know the level (of the labels) that introduces the difference. We have to create g\*(g-1)/2 intervals.

```{r}
group <- "tec"
subset_data <- subset(pisa_data,select = c(included_variables[[group]],"CNT"))
countries <- unique(pisa_data$CNT)

#datasets dimension
n <- dim(pisa_data)[1]
p <- length(included_variables[[group]]) 
g <- length(unique(pisa_data$CNT))

num_groups <- count(subset_data,CNT)

#parameters to compute CI
alpha <- 0.05
k <- p*g*(g-1)/2
qT <- qt(1-alpha/(2*k), n-g)
#computing the pooled estimate of the covariance matrix
W <- summary.manova(fit[[group]])$SS$Residuals

#means
mean_vector <- sapply(select_if(subset_data,is.numeric), mean)
mean_vector <- as.array(mean_vector)

#means by groups
group_means <- aggregate(select_if(subset_data,is.numeric), by=list(subset_data$CNT), FUN=mean)
group_means$Group.1 <- NULL

#computing confidence intervals
inf <- array(0,dim=c(g-1,g,p))
sup <- array(0,dim=c(g-1,g,p))
mean_diff <- array(0,dim=c(g-1,g,p))

#filling the confidence intervals
#i,j: iterates over the countries to compare them
#k: iterates over the variable
for(i in 1:(g-1)){
  for(j in (i+1):g){
    for(k in 1:p){
      #difference of means
      mean_diff[i,j,k] <- group_means[i,k]-group_means[j,k]
      #lower bound
      inf[i,j,k] <- (group_means[i,k]-group_means[j,k])-
                  qT * sqrt(diag(W)[k]/(n-g) * (1/num_groups[i,2]+1/num_groups[j,2]))
      #upper bound
      sup[i,j,k] <- (group_means[i,k]-group_means[j,k])+
                  qT * sqrt(diag(W)[k]/(n-g) * (1/num_groups[i,2]+1/num_groups[j,2]))
    }
  }
}
```

## Visualization

```{r}
ylim <- 1.5
cnt_comb <- combn(unique(countries),2)
diff_index <- paste(cnt_comb[1,],"vs",cnt_comb[2,]) 

legend_colors <- colorRampPalette(rainbow(length(countries)))(length(countries))

for(k in 1:p){
  plot(c(1,g*(g-1)/2),ylim=c(-ylim,ylim), xlim=c(c(1,g*(g-1)/2)), pch='',
        xlab='', ylab=paste('CI',k), 
        main=paste('CI',colnames(subset_data)[k]),xaxt='n')
  for(i in 1:(g-1)){
    for(j in (i+1):g){
      ind <- (i-1)*g-i*(i-1)/2+(j-i)
      #plotting the mean
      points(ind, mean_diff[i,j,k], pch=16); 
      #plotting inf and sup
      lines (c(ind,ind), c(inf[i,j,k],sup[i,j,k])); 
      points(ind, inf[i,j,k], col=rainbow(g)[i], pch=16); 
      points(ind, sup[i,j,k], col=rainbow(g)[j], pch=16); 
    }
  }
  abline(h=0)
  axis(1, at=1:(g*(g-1)/2), labels=diff_index[1:(g*(g-1)/2)],las=2,lwd=0.001,cex.axis = 0.7)
  legend("bottomright",legend = countries, col = legend_colors,cex = 0.5, pch = rep(1,length(countries)))
}
```
