---
title: "PERMANOVA"
output: 
date: "2023-04-16"
---

# NOTES

-   Permutational MANOVA tests are computationally expensive: we need a much smaller dataset than the one we're working on. We may reduce our dataset grouping by schools. Considering less countries does not work: still too many rows for a permutational test
-   I'm exploring options such as kruskal-wallis tests to see if I can make it work

# REFERENCES

-   theoretical introduction to PERMANOVA with a tutorial on how to implement it <https://uw.pressbooks.pub/appliedmultivariatestatistics/chapter/permanova/>
-   options to work with a large dataset (parallel computing and other hpc techniques implementation in r)(could be intersting but difficult to implement it): <https://cran.r-project.org/web/views/HighPerformanceComputing.html>

# PRELIMINARY STUFF

```{r}
#directories
root_proj_dir = "../../"
dataset_path = paste(root_proj_dir,"/data/pisa-woNA_school_final.csv",sep="")
include_path = paste(root_proj_dir,"/src/include/Utilities.R",sep="")
#including utilities
source(include_path)
#importing the dataset
pisa_data <- read.csv(file=dataset_path)
```

```{r}
#some adjustments on the data
pisa_data$X <- NULL
pisa_data$schID <- NULL
pisa_data$CNT <- as.factor(pisa_data$CNT)
```

# NON-PARAMETRIC MANOVA

## DISTANCE-BASED

Computationally expensive: need a really small dataset. I keep them here because they could be intersting but I set an option at chunk level not to run them \`{r eval=FALSE, include=FALSE}\`

### ADONIS2

```{r eval=FALSE, include=FALSE}
library(parallel)
library(vegan)

#try with smaller datasets
small_pisa <- pisa_data %>% filter(CNT %in% c("FRA","SWE"))

result <- adonis2(formula = small_pisa[,c(stu_ICT,"RATCMP1")]~CNT, 
                  data = small_pisa, 
                  method = "euclidean",
                  parallel = detectCores())
```

### PERMANOVA

```{r eval=FALSE, include=FALSE}
library(PERMANOVA)
small_pisa <- pisa_data %>% filter(CNT %in% c("FRA","IRL"))

#standardize data
pisa_std <- IniTransform(select_if(small_pisa,is.numeric)) 

#compute the matrix of distances
D <- DistContinuous(pisa_std) 
#adding grouping columns to the transformed data
pisa_std["CNT"] <- as.factor(small_pisa$CNT) 

#PERMANOVA
perm_pisa <- PERMANOVA(D,pisa_std$CNT)
```

## KRUSKAL-WALLIS

Another non-parametric test, less computationally expensive

```{r}
library(coin)
help("kruskal.test")
pisa_data$CNT <- as.factor(pisa_data$CNT)

#constructing the matrix to be fed to kruskal test
kruskal_test(ICTHOME ~ CNT, data = pisa_data)
```

Still doesn't work

```{r}
clt <- c("JOYREAD","CULTPOSS","HEDRES","SCREADCOMP","LMINS","MMINS")
#building the formula object
formula_str <- paste(clt, collapse=",")
formula_str <- paste("cbind(",formula_str,")",sep = "")
formula_str <- paste(formula_str,"CNT",sep = "~")

#performing the test
kruskal_test(formula = as.formula(formula_str),data = pisa_data)
```

Still need to be fixed

```{r eval=FALSE, include=FALSE}
library(npmv)
help(npmv)

quartz()
nonpartest(ICTCLASS|ICTHOME|ENTUSE|ICTHOME~CNT,data = pisa_data)
```
