---
title: "R Notebook"
# output: html_notebook
editor_options: 
  chunk_output_type: inline
---

```{r}
library(car)
library(MVN)
library(mvtnorm)
library(mvnormtest)

library(ggplot2)
library(grid)
library(sf)
library(dplyr)
library(rnaturalearth)

library(sp)

library(ggrepel)
library(tmap)
library(knitr)

#directories
root_proj_dir = "../../"
dataset_path = paste(root_proj_dir,"/data/pisa_scores_final.csv",sep="")
include_path = paste(root_proj_dir,"/src/include/Utilities.R",sep="")
#including utilities
source(include_path)
#importing the dataset
data <- read.csv(file=dataset_path)

head(data)
```

# **0** DATA PREP
```{r}
data$X <- NULL
data$SCHLTYPE <- as.factor(data$SCHLTYPE)
data$CNT <- as.factor(data$CNT)

IM_PUBLIC = rep(0,dim(data)[1])
IM_PUBLIC [which(data$SCHLTYPE=="Public")] = 1
data$IM_PUBLIC = as.factor(IM_PUBLIC)
data$SCHLTYPE <- NULL
head(data)
linear_model_vars <- readLines("../../data/non csv/lm_social_vars.txt")
```


## **2.** Exploration
```{r, warning=FALSE, message=FALSE}
p =length(linear_model_vars)
add_well_being = 1
if(add_well_being){
  linear_model_vars = c(linear_model_vars,"Social.well.being","Psychological.well.being")
  p=p+2
}

measures <- data[,linear_model_vars]

Label_1 <- as.factor(data[,"CNT"])
factor1 <- levels(Label_1)
g <- length(factor1)

Label_2 <- as.factor(data[,"IM_PUBLIC"])
factor2 <- levels(Label_2)
b <- length(factor2)

head(data)
n <- dim(data)[1]
n
```

## Data prep
```{r, warning=FALSE, message=FALSE}
# Re-name
lab_1_index=24
lab_2_index=27
colnames(data)[lab_1_index]='label_1'
colnames(data)[lab_2_index]='label_2'
head(data)

index_cnt = grep("label_1",colnames(data))
levels(data[,index_cnt])=levels(data$label_1)
cat_var1 <- data[,index_cnt]
index_public= grep("label_2",colnames(data))
levels(data[,index_public])=levels(data$label_2)
cat_var2 <- data[,index_public]
Label_1_2 <- interaction(cat_var1, cat_var2, sep = "-")
Label_1_2 = as.factor(Label_1_2)



## Graphical exploration of the data
for ( i in 1:p){
  # Effect of the treatments + their interaction on the i-th variable
  layout(matrix(c(1,1,2,3), 2, byrow=T))
  boxplot(measures[,i]~Label_1_2, main=paste('With Interac. lab_1+lab_2','variable:',linear_model_vars[i]),
  ylab='Tr', col='grey95')
  boxplot(measures[,i]~Label_1, main='Only lab_1', ylab='Tr',col=c('red','blue'))
  boxplot(measures[,i]~Label_2, main='Only lab_2', ylab='Tr',col=c('forestgreen','gold'))
}
```

## **3.** Model Assumptions
```{r, warning=FALSE, message=FALSE}
# Note that: if we have to remove rows -> one at the time
# Assumptions

# 1) Gaussianity
treat = levels(Label_1_2)
treat
gb = length(treat)
gb
```


```{r, warning=FALSE, message=FALSE}
for (i in 1:gb){
  if(dim(measures[Label_1_2==treat[i],])[1]>7){
    print(paste("On", dim(measures[Label_1_2==treat[i],])[1],"samples:",
                "pvalue for normality",treat[i],"=",
              mvn(measures[Label_1_2==treat[i],])$multivariateNormality$"p value"))
  }else
    print(paste("On", dim(measures[Label_1_2==treat[i],])[1],"samples:",
                "not enough samples for checking",treat[i]))
}
```


```{r, warning=FALSE, message=FALSE}
library(MVN)
library(car)
library(heplots)
# 2) homogeneity of the covariance (qualitatively)
S_list = list()
for ( i in 1:gb){
  S_list[i] = list(cov(measures[Label_1_2==levels(Label_1_2)[i], ]))
}

S_bind <- do.call(rbind, S_list)
```


```{r, warning=FALSE, message=FALSE}
summary(boxM(measures,Label_1_2))

cols = gray.colors(100)
for (i in 1:gb)
image(cov(measures[Label_1_2==treat[i],]),
      breaks = quantile(cov(measures[Label_1_2==treat[i],]),(0:100)/100),
      col=cols)
# 'nsomma
```
 
##**5.** Models 
### Complete model
 measure.ijk = mu + tau.i + beta.j + gamma.ij + eps.ijk eps.ijk ~ N(0, sigma^2)  
 i=1,2 (label_1)  
 j=1,2 (label_2)  
```{r, warning=FALSE, message=FALSE}
# what is the difference with Label_1:Label_2
fit2.int = manova(as.matrix(measures) ~ Label_1 + Label_2 + Label_1:Label_2)

summary(fit2.int, test="Wilks")
```

### Additive model
 measure.ijk = mu + tau.i + beta.j + eps.ijk eps.ijk ~ N(0, sigma^2)
 i=1,2 (label_1)
 j=1,2 (label_2)
```{r, warning=FALSE, message=FALSE}
fit2.ad = manova(as.matrix(measures) ~ Label_1 + Label_2)
summary(fit2.ad, test="Wilks")
```

## **6.** Which supplement is responsible?
```{r, warning=FALSE, message=FALSE}
# First of all:
#   Let's see on which variables the group has an effect.
#   Via ANOVA: for each feature we perform an ANOVA to see if the belonging to
#   the group has an effect on the mean of the variables.
summ = summary.aov(fit2.int)


# j is the label, k the variable
for( j in 1:2){
  pval = c()
  for ( k in 1:p)
    pval=c(pval,summ[[k]][j, 5])
  hist(log(pval))
}



# Comment
# Pr(>F) = p-value small -> the group has an influence on that X_k
# This analysis does NOT say either which groups differ nor which are the variables
# for which the groups differ.
```

## statistic preparation
```{r, warning=FALSE, message=FALSE}
W = summary.manova(fit2.int)$SS$Residuals
Gdl <- n-g-b+1
SSres <- t(fit2.ad$residuals) %*% fit2.ad$residuals / Gdl 

alpha = 0.1


# How many comparisons?
k = g*(g-1)/2*p + b*(b-1)/2*p
k

qT = qt(1 - alpha / (2 * k), Gdl)
ng = tapply(measures[,1],Label_1,length)
nb = tapply(measures[,1],Label_2,length)

mean_lab_1 <-list()
for(j in 1:p){
  mean_lab_1[j]<-list(tapply(measures[[j]],Label_1,mean))
}

mean_lab_2 <-list()
for(j in 1:p){
  mean_lab_2[j]<-list(tapply(measures[[j]],Label_2,mean))
}
```
# line plot of the mean
```{r}
for(i in 1:p){
  fdx <- vector(length = g)
  ordered_mean = mean_lab_1[[i]][order(mean_lab_1[[i]])]
  plot(1:g,ordered_mean,type="b",pch =16,main=colnames(measures)[i])
  for (j in 2:(g-1)) {
    fdx[j] <- (ordered_mean[j] - ordered_mean[j-1]) 
  }
  plot(1:g,fdx,type="b",pch =16,main=paste(colnames(measures)[j],"discrete derivative"))
  
}


  # Iterate through the values using the forward differencing method

```


## BONFERRONI CI
```{r, warning=FALSE, message=FALSE}
k = p*g*(g-1)/2
n = dim(data)[1]

CI = list()
confronti=NULL
for (col in 1:p){
  CI_temp = NULL
  for (i in 1:(g-1)){
    for (j in (i+1):g){
      if(col==1){
         confronti = c(confronti,(paste(factor1[i],"-",factor1[j])))
      }
     
      inf = mean_lab_1[[col]][i]-mean_lab_1[[col]][j]-qT*sqrt(diag(W)[col]/(n-g) * (1/ng[i]+1/ng[j]))
      mean= mean_lab_1[[col]][i]-mean_lab_1[[col]][j]
      sup = mean_lab_1[[col]][i]-mean_lab_1[[col]][j]+qT*sqrt(diag(W)[col]/(n-g) * (1/ng[i]+1/ng[j]))
      INT = cbind("inf" = inf, "mean"=mean,"sup"=sup)
      #rownames(INT)=colnames(data)
      CI_temp = rbind(CI_temp,INT)
    }
  }
  rownames(CI_temp)=confronti
  CI[[colnames(measures)[col]]]=CI_temp
}
```



```{r}
gruoup_ordered_list = list()
livelli_fittizzi = c(rep("low",5),rep("medium",5),rep("high",4))
for(k in 1:p){
    group_ordered <- with(data,                      
                      reorder(data$label_1,
                              measures[,k],
                              FUN=median))
    boxplot(measures[,k]~group_ordered, main=paste(colnames(measures)[k],"ordered by median"),las=2,
          ylim=c(round(min(measures[,k])),
          round(max(measures[,k]))), col = rainbow(g))
    abline(h=0)
   

    
    ###### 
    #         plot(c(1,g*(g-1)/2),ylim=c(min(unlist(CI[[k]])),max(unlist(CI[[k]]))), xlim=c(c(1,g*(g-1)/2)),
    #      pch='',xlab='pairs treat', ylab=paste('CI tau',colnames(measures)[k]), 
    #      main=paste('CI tau',colnames(measures)[k]))
    # abline(h=0)
    # 
    # for(i in 1:(g-1)) {
    #   for(j in (i+1):g) {
    #     ind <- (i-1)*g-i*(i-1)/2+(j-i)
    #   
    #     if(CI[[k]][ind,1]*CI[[k]][ind,3]>0){
    #         lines (c(ind,ind), c(CI[[k]][ind,1],CI[[k]][ind,3])); 
    #         points(ind, CI[[k]][ind,2], col="black", pch=16); 
    #         points(ind, CI[[k]][ind,1], col=rainbow(g)[j], pch=16); 
    #         points(ind, CI[[k]][ind,3], col=rainbow(g)[i], pch=16); 
    #     }
    #  }
    # }
######

    plot(c(1,g),ylim=c(min(unlist(CI[[k]])),max(unlist(CI[[k]]))),
         xlim=c(1,g),
         pch='',xlab='pairs treat', ylab=paste('CI tau',colnames(measures)[k]), 
         main=paste('CI tau',colnames(measures)[k]))
    abline(h=0)
    
    factor_ord = order(mean_lab_1[[p]])
    country_ordered = factor1[factor_ord[1]]
    livelli_final = 0
    lev=0
    for(i in 2:g){
      
      if(factor1[factor_ord[i-1]]<factor1[factor_ord[i]]){
        row = paste(factor1[factor_ord[i-1]],"-",factor1[factor_ord[i]])
      }
      else{
        row = paste(factor1[factor_ord[i]],"-",factor1[factor_ord[i-1]])
      }
      if(CI[[k]][row,1]*CI[[k]][row,3]>0 && lev<8){
        lev=lev+1
        lines (c(i,i), c(CI[[k]][row,1],CI[[k]][row,3])); 
        points(i, CI[[k]][row,2], col="black", pch=16); 
        points(i, CI[[k]][row,1], col=rainbow(g)[i-1], pch=16); 
        points(i, CI[[k]][row,3], col=rainbow(g)[i], pch=16); 
        
      }

      
      livelli_final = c(livelli_final,lev)
      country_ordered = c(country_ordered,factor1[factor_ord[i]])
    }
    print(livelli_final)
    
    column = colnames(measures)[k]
    gruoup_ordered_list[[column]] = data.frame(country_ordered,livelli_final)
    gruoup_ordered_list[[column]]$livelli_final=as.factor(
      gruoup_ordered_list[[column]]$livelli_final)
    

}
  
```

# which variables produce the most difference
```{r}
counter=list()
for ( col in 1:p){
  temp = CI[[colnames(measures)[col]]]
  count=0
  for(i in 1:(g*(g-1)/2)){
    if(temp[i,1]*temp[i,2]>0)
      count=count+1
  }
  counter[[colnames(measures)[col]]]=count
}
counter
```





# IM_PUBLIC
```{r, warning=FALSE, message=FALSE}
k = p*b*(b-1)/2
n = dim(data)[1]

CI_2 = list()
confronti=NULL
for (col in 1:p){
  CI_temp = NULL
  for (i in 1:(b-1)){
    for (j in (i+1):b){
      if(col==1){
         confronti = c(confronti,(paste(factor2[i],"-",factor2[j])))
      }
     
      inf = mean_lab_2[[col]][i]-mean_lab_2[[col]][j]-qT*sqrt(diag(W)[col]/(n-g) * (1/nb[i]+1/nb[j]))
      sup = mean_lab_2[[col]][i]-mean_lab_2[[col]][j]+qT*sqrt(diag(W)[col]/(n-g) * (1/nb[i]+1/nb[j]))
      INT = cbind("inf" = inf, "sup"=sup)
      #rownames(INT)=colnames(data)
      CI_temp = rbind(CI_temp,INT)
    }
  }
  rownames(CI_temp)=confronti
  CI_2[[colnames(measures)[col]]]=CI_temp
}
```
# which variables produce the most difference
```{r}
counter_2=list()
for ( col in 1:p){
  temp = CI[[colnames(measures)[col]]]
  count=0
  for(i in 1:(b*(b-1)/2)){
    counter_2[[colnames(measures)[col]]]=0
    if(temp[i,1]*temp[i,2]>0){
    if(temp[i,1]>0){
      counter_2[[colnames(measures)[col]]]=1
    }else{
      counter_2[[colnames(measures)[col]]]=2
    }
      
  }
  
  }
}

counter_2
```

# DATA PREP FOR MAPS
```{r}
all_countries <- c("AUT", "BEL", "BGR", "CYP", "CZE", "DEU", "DNK", "ESP", "EST", "FIN", "FRA", "GRC", "HRV", "HUN", "IRL", "ITA", "LTU", "LUX", "LVA", "MLT", "NLD", "POL", "PRT", "ROU", "SVK", "SVN", "SWE")


europe_map <- ne_countries( scale=50,returnclass = 'sf',continent = "europe")
sf_use_s2(FALSE)
centroids <- st_centroid(europe_map)

europe_map <- cbind(europe_map, st_coordinates(st_centroid(europe_map$geometry)))
```

# CONTINUOUS MAPS
```{r}

for (k in 1:1){
    means = data.frame(mean_lab_1[[k]],names(mean_lab_1[[k]]))
    colnames(means)=c("mean_value","adm0_a3")
    for (country in all_countries) {
      # Check if the country is already in the dataset
      if (!(country %in% means$adm0_a3)) {
        # If not, add the country to the dataset with a value of 0
        means <- rbind(means, data.frame(adm0_a3 = country, mean_value = NA))
      }
    }

    europe_joined <- merge(europe_map, means, by = c("adm0_a3"))
    
    print(
      europe_joined %>% ggplot(aes(fill = mean_value)) +
      geom_sf(size = 0.2, color = "black") + # border line
     scale_fill_continuous(
      type = "viridis",
      name = "%change",                  # title of the legend
      na.value = "white",  # Specify the color for NA values
      guide = guide_colorbar(
        direction = "vertical",             # vertical colorbar
        title.position = "top",             # title displayed at the top
        label.position = "right",           # labels displayed at the right side
        barwidth = unit(0.4, "cm"),         # width of the colorbar
        barheight = unit(7, "cm"),          # height of the colorbar
        ticks = TRUE))+
      labs(
        title =colnames(measures)[k],
        subtitle = "MANOVA") +
      scale_x_continuous(limits = c(-10, 35)) +
      scale_y_continuous(limits = c(35, 68))+
      theme_void()+
      geom_text_repel(aes(x = X, y = Y, label = name),
                data = subset(europe_joined, !is.na(mean_value)),
                size = 3.5,                   # Increase the size of the labels
                colour = "black",          # Set the color of the labels
                fontface = "bold",          # Set the font weight to bold
                label.padding = unit(0.5, "lines"),   # Add padding around the label
                label.r = unit(0.2, "lines"),         # Set the radius of the label's rounded corners
                label.color = "black",                 # Set the color of the label's border
                label.size = 0.8,                      # Set the size of the label's border
                label.fill = "white"                    # Set the color of the label's background
                ) 
    
    )
}

```
#DISCRETE MAPS
```{r}
for (k in 1:3){
    means = gruoup_ordered_list[[k]]
    means = data.frame(gruoup_ordered_list[[k]][,2],gruoup_ordered_list[[k]][,1])
    colnames(means)=c("mean_value","adm0_a3")
    for (country in all_countries) {
      # Check if the country is already in the dataset
      if (!(country %in% means$adm0_a3)) {
        # If not, add the country to the dataset with a value of 0
        means <- rbind(means, data.frame(adm0_a3 = country, mean_value = NA))
      }
    }

    europe_joined <- merge(europe_map, means, by = c("adm0_a3"))
    
    print(
      europe_joined %>% ggplot(aes(fill = mean_value)) +
      geom_sf(size = 0.2, color = "black") + # border line
    scale_fill_manual(
      name = "legend", 
      breaks = levels(means$mean_value),
      ############### COLORI
      # values =brewer.pal(n = length(levels(means$mean_value)), name = "Blues"),
      values =colora(length(levels(means$mean_value))),
      ############### COLORI
      #labels = c("high", "medium", "low"),
      na.value = "white"  # Specify the color for NA values
      ) +
      labs(
        title =colnames(measures)[k],
        subtitle = "MANOVA") +
      scale_x_continuous(limits = c(-10, 30)) +
      scale_y_continuous(limits = c(35, 70))+
      theme_void()+
      geom_text_repel(aes(x = X, y = Y, label = name),
                data = subset(europe_joined, !is.na(mean_value)),
                size = 3.5,                   # Increase the size of the labels
                colour = "black",          # Set the color of the labels
                fontface = "bold",          # Set the font weight to bold
                ) 
    
    )
}

```


