---
title: "R Notebook"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

```{r}
library(car)
library(MVN)
library(mvtnorm)
library(mvnormtest)
#directories
root_proj_dir = "../../"
dataset_path = paste(root_proj_dir,"/data/pisa_scores_final.csv",sep="")
include_path = paste(root_proj_dir,"/src/include/Utilities.R",sep="")
#including utilities
source(include_path)
#importing the dataset
data <- read.csv(file=dataset_path)
head(data)
```

# **0** DATA PREP
```{r}
data$X <- NULL
data$SCHLTYPE <- as.factor(data$SCHLTYPE)
data$CNT <- as.factor(data$CNT)

IM_PUBLIC = rep(0,dim(data)[1])
IM_PUBLIC [which(data$SCHLTYPE=="Public")] = 1
data$IM_PUBLIC = as.factor(IM_PUBLIC)
data$SCHLTYPE <- NULL
head(data)
linear_model_vars <- readLines("../../data/lm_social_vars.txt")
```


## **2.** Exploration
```{r, warning=FALSE, message=FALSE}
p =length(linear_model_vars)
measures <- data[,linear_model_vars]

Label_1 <- as.factor(data[,"CNT"])
factor1 <- levels(Label_1)
g <- length(factor1)

Label_2 <- as.factor(data[,"IM_PUBLIC"])
factor2 <- levels(Label_2)
b <- length(factor2)

head(data)
n <- dim(data)[1]
n
```

## Data prep
```{r, warning=FALSE, message=FALSE}
# Re-name
lab_1_index=24
lab_2_index=27
colnames(data)[lab_1_index]='label_1'
colnames(data)[lab_2_index]='label_2'
head(data)

index_cnt = grep("label_1",colnames(data))
levels(data[,index_cnt])=levels(data$label_1)
cat_var1 <- data[,index_cnt]
index_public= grep("label_2",colnames(data))
levels(data[,index_public])=levels(data$label_2)
cat_var2 <- data[,index_public]
Label_1_2 <- interaction(cat_var1, cat_var2, sep = "-")
Label_1_2 = as.factor(Label_1_2)



## Graphical exploration of the data
for ( i in 1:p){
  # Effect of the treatments + their interaction on the i-th variable
  layout(matrix(c(1,1,2,3), 2, byrow=T))
  boxplot(measures[,i]~Label_1_2, main=paste('With Interac. lab_1+lab_2','variable:',linear_model_vars[i]),
  ylab='Tr', col='grey95')
  boxplot(measures[,i]~Label_1, main='Only lab_1', ylab='Tr',col=c('red','blue'))
  boxplot(measures[,i]~Label_2, main='Only lab_2', ylab='Tr',col=c('forestgreen','gold'))
}
```

## **3.** Model Assumptions
```{r, warning=FALSE, message=FALSE}
# Note that: if we have to remove rows -> one at the time

# Assumptions
# 1) Gaussianity
treat = levels(Label_1_2)
gb = length(treat)


# 2) homogeneity of the covariance (qualitatively)
S_list = list()
for ( i in 1:gb){
  S_list[i] = list(cov(measures[Label_1_2==levels(Label_1_2)[i], ]))
}

S_bind <- do.call(rbind, S_list)

```
 
##**5.** Models 
### Complete model
 measure.ijk = mu + tau.i + beta.j + gamma.ij + eps.ijk eps.ijk ~ N(0, sigma^2)  
 i=1,2 (label_1)  
 j=1,2 (label_2)  
```{r, warning=FALSE, message=FALSE}
# what is the difference with Label_1:Label_2
fit2.int = manova(as.matrix(measures) ~ Label_1 + Label_2 + Label_1:Label_2)

summary(fit2.int, test="Wilks")
```

### Additive model
 measure.ijk = mu + tau.i + beta.j + eps.ijk eps.ijk ~ N(0, sigma^2)
 i=1,2 (label_1)
 j=1,2 (label_2)
```{r, warning=FALSE, message=FALSE}
fit2.ad = manova(as.matrix(measures) ~ Label_1 + Label_2)
summary(fit2.ad, test="Wilks")
```

## **6.** Which supplement is responsible?
```{r, warning=FALSE, message=FALSE}
# First of all:
#   Let's see on which variables the group has an effect.
#   Via ANOVA: for each feature we perform an ANOVA to see if the belonging to
#   the group has an effect on the mean of the variables.
summ = summary.aov(fit2.int)


# j is the label, k the variable
for( j in 1:2){
  pval = c()
  for ( k in 1:p)
    pval=c(pval,summ[[k]][j, 5])
  hist(log(pval))
}



# Comment
# Pr(>F) = p-value small -> the group has an influence on that X_k
# This analysis does NOT say either which groups differ nor which are the variables
# for which the groups differ.
```

## **7.** Bonferroni: symmetric design assumption (n1 = n2 = n), change if needed
We want to know the level (of the labels) that introduces the difference.
##  SSres vs W??
```{r, warning=FALSE, message=FALSE}
W = summary.manova(fit2.int)$SS$Residuals
Gdl <- n-g-b+1
SSres <- t(fit2.ad$residuals) %*% fit2.ad$residuals / Gdl 

alpha = 0.1


# How many comparisons?
k = g*(g-1)/2*p + b*(b-1)/2*p
k

qT = qt(1 - alpha / (2 * k), Gdl)
ng = tapply(measures[,1],Label_1,length)
nb = tapply(measures[,1],Label_2,length)

mean_lab_1 <-list()
for(j in 1:p){
  mean_lab_1[j]<-list(tapply(measures[[j]],Label_1,mean))
}

mean_lab_2 <-list()
for(j in 1:p){
  mean_lab_2[j]<-list(tapply(measures[[j]],Label_2,mean))
}
```


```{r, warning=FALSE, message=FALSE}
k = p*g*(g-1)/2
n = dim(data)[1]

CI = list()
confronti=NULL
for (col in 1:p){
  CI_temp = NULL
  for (i in 1:(g-1)){
    for (j in (i+1):g){
      if(col==1){
         confronti = c(confronti,(paste(factor1[i],"-",factor1[j])))
      }
     
      inf = mean_lab_1[[col]][i]-mean_lab_1[[col]][j]-qT*sqrt(diag(W)[col]/(n-g) * (1/ng[i]+1/ng[j]))
      mean= mean_lab_1[[col]][i]-mean_lab_1[[col]][j]
      sup = mean_lab_1[[col]][i]-mean_lab_1[[col]][j]+qT*sqrt(diag(W)[col]/(n-g) * (1/ng[i]+1/ng[j]))
      INT = cbind("inf" = inf, "mean"=mean,"sup"=sup)
      #rownames(INT)=colnames(data)
      CI_temp = rbind(CI_temp,INT)
    }
  }
  rownames(CI_temp)=confronti
  CI[[colnames(measures)[col]]]=CI_temp
}
```
```{r}
for(j in 1:p){
  boxplot(measures[,j]~data$label_1, main=paste(colnames(measures)[j]),
          ylim=c(round(min(measures[,j])),
          round(max(measures[,j]))), col = rainbow(g))
   abline(h=0)
}
# Order boxes by median




### cambia nome titolo
for(k in 1:p){
    group_ordered <- with(data,                      
                      reorder(data$label_1,
                              measures[,k],
                              median))
  boxplot(measures[,k]~group_ordered, main=paste(colnames(measures)[k]),las=2,
          ylim=c(round(min(measures[,k])),
          round(max(measures[,k]))), col = rainbow(g))
   abline(h=0)
   
   
  plot(c(1,g*(g-1)/2),ylim=c(min(unlist(CI[[k]])),max(unlist(CI[[k]]))), xlim=c(c(1,g*(g-1)/2)), pch='',
       xlab='pairs treat', ylab=paste('CI tau',colnames(measures)[k]), main=paste('CI tau',colnames(measures)[k]))
    abline(h=0)
  
 for(i in 1:(g-1)) {
  for(j in (i+1):g) {
    ind <- (i-1)*g-i*(i-1)/2+(j-i)

    if(CI[[k]][ind,1]*CI[[k]][ind,3]>0){
        lines (c(ind,ind), c(CI[[k]][ind,1],CI[[k]][ind,3])); 
        points(ind, CI[[k]][ind,2], col="black", pch=16); 
        points(ind, CI[[k]][ind,1], col=rainbow(g)[j], pch=16); 
        points(ind, CI[[k]][ind,3], col=rainbow(g)[i], pch=16); 
    }

   }
  }
}
  
```

# which variables produce the most difference
```{r}
counter=list()
for ( col in 1:p){
  temp = CI[[colnames(measures)[col]]]
  count=0
  for(i in 1:(g*(g-1)/2)){
    if(temp[i,1]*temp[i,2]>0)
      count=count+1
  }
  counter[[colnames(measures)[col]]]=count
}
counter
```


# IM_PUBLIC
```{r, warning=FALSE, message=FALSE}
k = p*b*(b-1)/2
n = dim(data)[1]

CI_2 = list()
confronti=NULL
for (col in 1:p){
  CI_temp = NULL
  for (i in 1:(b-1)){
    for (j in (i+1):b){
      if(col==1){
         confronti = c(confronti,(paste(factor2[i],"-",factor2[j])))
      }
     
      inf = mean_lab_2[[col]][i]-mean_lab_2[[col]][j]-qT*sqrt(diag(W)[col]/(n-g) * (1/nb[i]+1/nb[j]))
      sup = mean_lab_2[[col]][i]-mean_lab_2[[col]][j]+qT*sqrt(diag(W)[col]/(n-g) * (1/nb[i]+1/nb[j]))
      INT = cbind("inf" = inf, "sup"=sup)
      #rownames(INT)=colnames(data)
      CI_temp = rbind(CI_temp,INT)
    }
  }
  rownames(CI_temp)=confronti
  CI_2[[colnames(measures)[col]]]=CI_temp
}
```
# which variables produce the most difference
```{r}
counter_2=list()
for ( col in 1:p){
  temp = CI[[colnames(measures)[col]]]
  count=0
  for(i in 1:(b*(b-1)/2)){
    if(temp[i,1]*temp[i,2]>0)
      count=count+1
  }
  counter_2[[colnames(measures)[col]]]=count
}
counter_2
```