---
title: "R Notebook"
# output: html_notebook
editor_options:
  chunk_output_type: inline
---


```{r}
rm( list = ls() )

graphics.off() # chiude tutti i device grafici
cat("\014") #pulisci console
library(mvtnorm)
library(MASS)
library(car)
library(rgl)
library(dplyr)
library(mvtnorm)

library(glmnet)

library(leaps)
library(ISLR)
```

```{r}
root_proj_dir = "../../"
dataset_path = paste(root_proj_dir,"data/pisa_scores_final.csv",sep="")
include_path = paste(root_proj_dir,"src/include/Utilities.R",sep="")
#INCLUDING UTILITIES
source(include_path)
#IMPORTING THE DATASET
data <- read.csv(file=dataset_path)
head(data)
```

# [0] DATA PREP
```{r}
data$X <- NULL
data$SCHLTYPE <- as.factor(data$SCHLTYPE)
data$CNT <- as.factor(data$CNT)

IM_PUBLIC = rep(0,dim(data)[1])
IM_PUBLIC [which(data$SCHLTYPE=="Public")] = 1
data$IM_PUBLIC = as.factor(IM_PUBLIC)
data$SCHLTYPE <- NULL
dim(data)
```

# scaling the data
```{r}
data_scaled = as.data.frame(scale(select_if(data,is.numeric)),data$CNT,data$IM_PUBLIC)
dim(data_scaled)
data = cbind("CNT" =data$CNT,"IM_PUBLIC"=data$IM_PUBLIC,data_scaled)
dim(data)
rownames(data)=NULL
```




# [1] Modello 1-Social.well.being
```{r}
FORMULA_COMPLETE_SOCIAL =formula(Social.well.being~.-Psychological.well.being-IM_PUBLIC)
fit_social = lm(FORMULA_COMPLETE_SOCIAL,data=data)
summary(fit_social)
b=coefficients(fit_social)
nv_max = length(b)-1
names(b)
```

# Modello 2-Psychological.well.being
```{r}
FORMULA_COMPLETE_PSICO = formula(Psychological.well.being~.-Social.well.being-IM_PUBLIC)
fit_psico = lm(FORMULA_COMPLETE_PSICO,data=data)
summary(fit_psico)
b=coefficients(fit_psico)
nv_max = length(b)-1
names(b)
```
#SOCIAL
```{r}
fm = fit_social
Y=data$Social.well.being
FORMULA_LM = FORMULA_COMPLETE_SOCIAL
```


# [2] Variable selection
## Best Subset Selection (exhaustive search)
```{r, warning=FALSE, message=FALSE}
regfit.full = regsubsets(FORMULA_LM, data=data, nvmax=nv_max)
reg.summary = summary(regfit.full)

# Which one we choose:
#reg.summary$which
# Plots
plot(reg.summary$adjr2, xlab="Number of Variables", ylab="Adjusted RSq", type="b")


# We want the model with max r.adj^2 so we extract the coefficients of that model
# Note: ind = how many coefficients has the model
ind = which.max(reg.summary$adjr2)
ind
coef(regfit.full, ind)
```



## Comparison- they are all the same
```{r, warning=FALSE, message=FALSE}
max_vars = 29
full = names(coef(regfit.full,max_vars)) # Exhaustive search
full

vars_chosen <- full[14:max_vars]


```


## **6.** Lasso Regression
```{r, warning=FALSE, message=FALSE}
x = model.matrix(FORMULA_LM,data=data)[,-1] # matrix of predictors
y = Y # vector of response
lambda.grid = 10^seq(4,-5,length=50)


# Lasso regression
fit.lasso = glmnet(x,y, lambda = lambda.grid, alpha=1) # alpha=1 -> lasso
plot(fit.lasso,xvar='lambda',label=TRUE, col = rainbow(dim(x)[2]))
legend('topright', dimnames(x)[[2]], col = rainbow(dim(x)[2]), lty=1, cex=1)


# Set lambda via CV
cv.lasso = cv.glmnet(x,y,alpha=1,nfolds=10,lambda=lambda.grid)
bestlam.lasso = cv.lasso$lambda.min
bestlam.lasso

plot(cv.lasso)
abline(v=log(bestlam.lasso), lty=1)

# Get the coefficients for the optimal lambda
coef.lasso.cv = predict(fit.lasso, s=bestlam.lasso, type = 'coefficients')
lasso_vars = coef.lasso.cv[which(abs(coef.lasso.cv)> 0.01),]
lasso_vars_names = names(lasso_vars)

plot(fit.lasso,xvar='lambda',label=TRUE, col = rainbow(dim(x)[2]))
abline(v=log(bestlam.lasso))
```


```{r}
final_vars = union(vars_chosen,lasso_vars_names[-1])

final_vars = vars_chosen
setdiff(colnames(data),final_vars)


final_vars_social <- final_vars
```


# PSYCHO
```{r}
fm = fit_psico
Y=data$Psychological.well.being
FORMULA_LM = FORMULA_COMPLETE_PSICO
```


# [2] Variable selection
## Best Subset Selection (exhaustive search)
```{r, warning=FALSE, message=FALSE}
regfit.full = regsubsets(FORMULA_LM, data=data, nvmax=nv_max)
reg.summary = summary(regfit.full)

# Which one we choose:
#reg.summary$which
# Plots
plot(reg.summary$adjr2, xlab="Number of Variables", ylab="Adjusted RSq", type="b")


# We want the model with max r.adj^2 so we extract the coefficients of that model
# Note: ind = how many coefficients has the model
ind = which.max(reg.summary$adjr2)
ind
coef(regfit.full, ind)
```



## Comparison- they are all the same
```{r, warning=FALSE, message=FALSE}
max_vars = 17
full = names(coef(regfit.full,max_vars))
full
vars_chosen <- full[2:max_vars]
```


## **6.** Lasso Regression
```{r, warning=FALSE, message=FALSE}
x = model.matrix(FORMULA_LM,data=data)[,-1] # matrix of predictors
y = Y # vector of response
lambda.grid = 10^seq(4,-5,length=50)


# Lasso regression
fit.lasso = glmnet(x,y, lambda = lambda.grid, alpha=1) # alpha=1 -> lasso
plot(fit.lasso,xvar='lambda',label=TRUE, col = rainbow(dim(x)[2]))
legend('topright', dimnames(x)[[2]], col = rainbow(dim(x)[2]), lty=1, cex=1)


# Set lambda via CV
cv.lasso = cv.glmnet(x,y,alpha=1,nfolds=10,lambda=lambda.grid)
bestlam.lasso = cv.lasso$lambda.min
bestlam.lasso

plot(cv.lasso)
abline(v=log(bestlam.lasso), lty=1)

# Get the coefficients for the optimal lambda
coef.lasso.cv = predict(fit.lasso, s=bestlam.lasso, type = 'coefficients')
lasso_vars = coef.lasso.cv[which(abs(coef.lasso.cv)> 0.01),]
lasso_vars_names = names(lasso_vars)

plot(fit.lasso,xvar='lambda',label=TRUE, col = rainbow(dim(x)[2]))
abline(v=log(bestlam.lasso))
```


```{r}
final_vars = union(vars_chosen,lasso_vars_names[-1])
setdiff(colnames(data),final_vars)

final_vars_psico <- final_vars

```



# LINEAR HYP
```{r}
linear_model_vars <- c(final_vars_social
                      ,"CNT"
                      )

vars = c(linear_model_vars)
FORMULA_SOCIAL <- paste(paste("Social.well.being","~"), paste(vars, collapse = "+"))
fit_social_lm = lm(FORMULA_SOCIAL,data=data)

summary(fit_social_lm)
```

## SOCIAL 
```{r}
b=coefficients(fit_social_lm)

vars_to_discard <- c("RATCMP1",
                     "JOYREAD",
                     "PROAT6",
                     "EDUSHORT")

vars_to_discard <- c(
        "Teachers..degree",
        "ESCS" ,
        "STUBEHA", 
        "PROAT6" ,
        "CLSIZE" ,
        "STAFFSHORT", 
        "CREACTIV" 
)


check_beta <- which(names(b)%in%vars_to_discard)
C <- c()
for( i in 1: length(check_beta)){
  vect <- rep(0,fit_social_lm$rank)
  vect[check_beta[i]] <- 1
  C <- rbind(C,vect)
}
linearHypothesis(fit_social_lm, C, rep(0,length(check_beta)))

```

```{r}
vars_final = vars[!(vars%in%c(vars_to_discard,"CNT"))]
FORMULA_SOCIAL <- paste(paste("Social.well.being","~"), paste(c(vars_final,"CNT"), collapse = "+"))

writeLines(vars_final,"../../data/non csv/lm_social_vars.txt")

fit_social_lm = lm(FORMULA_SOCIAL,data=data)
summary(fit_social_lm)
```

## PSYCHO

```{r}
linear_model_vars <-c(final_vars_psico
                      ,"CNT"
                      )

vars = c(linear_model_vars)
FORMULA_PSYCH <- paste(paste("Psychological.well.being","~"), paste(vars, collapse = "+"))
fit_psych_lm = lm(FORMULA_PSYCH,data=data)
```

```{r}
summary(fit_psych_lm)
```

```{r}
b=coefficients(fit_psych_lm)
vars_to_discard <- c("RATCMP1",
                     "PROAT6",
                     "EDUSHORT",
                     "JOYREAD",
                     "TEACHBEHA",
                     "STUBEHA")

vars_to_discard = c("Use.of.ICT",
                    "ESCS",
                    "RATCMP1",
                    "ICTRES",
                    "PROAT6",
                    "CLSIZE",
                    "EDUSHORT",
                    "PV1READ",
                    "JOYREAD",
                    "TEACHBEHA",
                    "CREACTIV",
                    "STUBEHA"
                    )




check_beta <- which(names(b)%in%vars_to_discard)
C <- c()
for( i in 1: length(check_beta)){
  vect <- rep(0,fit_psych_lm$rank)
  vect[check_beta[i]] <- 1
  C <- rbind(C,vect)
}
linearHypothesis(fit_psych_lm, C, rep(0,length(check_beta)))

```

```{r}
vars_final = vars[!(vars%in%c(vars_to_discard,"CNT"))]
FORMULA_PSYCH <- paste(paste("Psychological.well.being","~"), paste(c(vars_final,"CNT"), collapse = "+"))

writeLines(vars_final, "../../data/non csv/lm_psico_vars.txt")

fit_psych_lm = lm(FORMULA_PSYCH,data=data)
summary(fit_psych_lm)
```
# Grouped variables remaining
## social
```{r}
linear_model_vars <- readLines("../../data/non csv/lm_social_vars.txt")

filter <- function(category,var_lm) {
  risultato <- c()
  
  for (elemento in category) {
    if (elemento %in% var_lm) {
      risultato <- c(risultato, elemento)
    }
  }
  
  return(risultato)
}


categories_variables_filtered=list()

for(cat in cat_var_names ){
  categories_variables_filtered[[cat]]=filter(categories_variables[[cat]],linear_model_vars)
}

categories_variables_filtered
```
## psico
```{r}
linear_model_vars <- readLines("../../data/non csv/lm_psico_vars.txt")

filter <- function(category,var_lm) {
  risultato <- c()
  
  for (elemento in category) {
    if (elemento %in% var_lm) {
      risultato <- c(risultato, elemento)
    }
  }
  
  return(risultato)
}


categories_variables_filtered=list()

for(cat in cat_var_names ){
  categories_variables_filtered[[cat]]=filter(categories_variables[[cat]],linear_model_vars)
}

categories_variables_filtered
```
