---
title: "exploring data (after PCA)"
output: 
editor_options: 
  chunk_output_type: inline
---

# EXPLORATORY ANALYSIS: PCA SCORES

In this notebook I collect some useful plots made across the other notebooks that help better understanding the dataset grouped by schools

## PRELIMINARY STUFF

```{r, setup}
#DIRECTORIES
root_proj_dir = "../../"
dataset_path = paste(root_proj_dir,"data/df_pca_scores.csv",sep="")
include_path = paste(root_proj_dir,"src/include/Utilities.R",sep="")
#INCLUDING UTILITIES
source(include_path)
#IMPORTING THE DATASET
pisa_data <- read.csv(file=dataset_path)
head(pisa_data)
```

```{r}
all_vars = colnames(pisa_data)

tech <- colnames(pisa_data[grep("tec", all_vars)])
tech
psic <- colnames(pisa_data[grep("psi", all_vars)])
psic
culture <- colnames(pisa_data[grep("clt", all_vars)])
culture
family <- colnames(pisa_data[grep("fam", all_vars)])
family
teach <- colnames(pisa_data[grep("tch", all_vars)])
teach
school <- colnames(pisa_data[grep("sch", all_vars)])
school
categories_variables = list(tech,psic,culture,family,teach,school)
cat_var_names = c("tec","psi","clt","fam","tch","sch")
```

```{r}
#some adjustments on the data
pisa_data$X <- NULL
pisa_data$CNT <- as.factor(pisa_data$CNT)
```

# CORRELATION

```{r}
library(GGally)
library(ggplot2)
library(dplyr)

ggcorr(select_if(pisa_data,is.numeric),label=TRUE,label_size = 2)
```

# NUMEROSITY BY COUNTRY

```{r}
ggplot(pisa_data, aes(x = CNT)) +
  geom_bar()
```

# BOXPLOTS

Boxplots for the full dataset

```{r}
boxplot(select_if(pisa_data,is.numeric),las=2,cex.axis=0.8)
```

# HISTOGRAMS: DISTRIBUTIONS FOR EACH VARIABLE

```{r}
countries = unique(pisa_data$CNT)
shap <-c()

for(c in countries){
  for(i in 2:dim(pisa_data)[2]){
  hist(pisa_data[which(pisa_data$CNT==c),i],main = paste(colnames(pisa_data)[i],c))
  shap <- c(shap,shapiro.test(pisa_data[which(pisa_data$CNT==c),i])$p.value)
}
}
plot(seq(1,(dim(pisa_data)[2]-1)*length(countries)),shap)
hist(shap)
```

# Scatterplot for each variable colored by countrys

```{r}
# output on console
for (categ in categories_variables){
  x11()
  par(mfrow = c(length(categ),1))
  
  data_categ = pisa_data[,c("CNT",categ)]
  data_categ$CNT <-as.factor(data_categ$CNT)

  for(i in 2:dim(data_categ)[2]){
   # boxplot(data_categ[,i],main = paste(colnames(data_categ)[i]),col=data_categ$CNT,pch=16)
      boxplot(data_categ[,i]~data_categ$CNT, main= paste(colnames(data_categ)[i]),las=2,
              col=rainbow(16))
  }
}
```

```{r}
pisa_data$CNT = as.factor(pisa_data$CNT)
plot(pisa_data[,school], col=pisa_data$CNT,pch=16)
legend("bottomright",legend=unique(pisa_data$CNT),col=1:length(pisa_data$CNT),pch=16)
```

```{r}
for (col in colnames(pisa_data)){
  plot(pisa_data[,col],col=1:length(pisa_data$CNT),pch=16,main = col)
}
```

