---
title: "R Notebook"
output: 
editor_options: 
  chunk_output_type: inline
---

# EXPLORATORY ANALYSIS: SCHOOL DATASET

In this notebook I collect some useful plots made across the other notebooks that help better understanding the dataset grouped by schools

## PRELIMINARY STUFF

```{r}
#libraries
library(GGally) #for ggcorr
library(ggplot2)
library(dplyr)
library(rlang) #to select variables to be plugged into a formula object
```

```{r, setup}
#DIRECTORIES
root_proj_dir = "../../"
dataset_path = paste(root_proj_dir,"data/pisa-woNA_school_final.csv",sep="")
include_path = paste(root_proj_dir,"src/include/Utilities.R",sep="")
#INCLUDING UTILITIES
source(include_path)
#IMPORTING THE DATASET
pisa_data <- read.csv(file=dataset_path)
```

```{r}
categories_variables = list(tec,psi,clt,fam,tch,sch)
cat_var_names = c("tec","psi","clt","fam","tch","sch")
```

```{r}
#some adjustments on the data
pisa_data$X <- NULL
pisa_data$schID <- as.factor(pisa_data$schID)
pisa_data$CNT <- as.factor(pisa_data$CNT)
```

# CORRELATION

```{r}
ggcorr(select_if(pisa_data,is.numeric),label=TRUE,label_size = 2)

#too many vars: unreadable
#ggpairs(select_if(pisa_data,is.numeric))
```

# NUMEROSITY BY COUNTRY

```{r}
ggplot(pisa_data, aes(x = CNT)) +
  geom_bar()
```

# BOXPLOTS

Boxplots for the full dataset

```{r}
#Here we see that the variability in LMINS and MMINS is very high
boxplot(select_if(pisa_data,is.numeric),las=2,cex.axis=0.8)
#removing the variables with the highest values
boxplot(select_if(subset(pisa_data,select=-c(LMINS,MMINS)),is.numeric),las=2,cex.axis=0.8)
boxplot(select_if(subset(pisa_data,select=-c(LMINS,MMINS,BFMJ2,BMMJ1,HISEI)),is.numeric),las=2,cex.axis=0.8)
boxplot(select_if(subset(pisa_data,select=-c(LMINS,MMINS,BFMJ2,BMMJ1,HISEI,ICTHOME,HISCED,ICTSCH,RATCMP1)),is.numeric),las=2,cex.axis=0.8)
```

Boxplots by country

```{r}
data <- subset(pisa_data,select=-c(LMINS,MMINS,BFMJ2,BMMJ1,HISEI,ICTHOME,HISCED,ICTSCH,RATCMP1,schID))
#by country for each variable
for(variable in colnames(data)){
  if(variable != "CNT"){
    formula_str <- paste(variable,"CNT", sep = "~")
    boxplot(formula = as.formula(formula_str), data = data, las=2,cex.axis=0.8, main = variable)
  }
}
```

# HISTOGRAMS: DISTRIBUTION FOR EACH VARIABLE

Full dataset:

```{r}
for(variable in colnames(pisa_data)){
    if(variable != "schID" && variable != "CNT"){
      hist(pisa_data[,variable],main = variable)
    }
}
```

By country:

```{r,fig.show='hide'}
countries = unique(pisa_data$CNT)
#plotting the histograms
for(cntry in countries){
  for(variable in colnames(pisa_data)){
    if(variable != "schID" && variable != "CNT"){
      hist(pisa_data[which(pisa_data$CNT==cntry),variable],main = paste(variable,cntry))
    }
  }
}
```

# VARIANCES

To see which variables have the higher variance

fig \<- ggplot(data = as.data.frame(diag(S)), aes(x=)) + geom_bar() + labs(title = "Variance") + theme(axis.text.x = element_text(angle = 45, hjust = 1))

fig

```{r}
#computing the variances
variances <- apply(select_if(pisa_data,is.numeric), 2, var)
var_df <- data.frame(variable = names(variances), variance = variances)

#plotting
ggplot(data = var_df, aes(x = variable, y = variance)) + 
  geom_bar(stat = "identity", fill = "steelblue") + 
  labs(title = "Variances of Dataframe") +
  xlab("Variable") +
  ylab("Variance") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

#removing the variables with the highest variance to see the variability of the others
high_variance_vars = c("LMINS","MMINS","BFMJ2","BMMJ1","HISEI","ICTHOME","HISCED","ICTSCH","RATCMP1")

ggplot(data = var_df[!(var_df$variable %in% high_variance_vars),], aes(x = variable, y = variance)) + 
  geom_bar(stat = "identity", fill = "steelblue") + 
  labs(title = "Variances of Dataframe") +
  xlab("Variable") +
  ylab("Variance") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
