# Indeces to measure Well-Being

## References

-   notebook on Factor Analysis in this folder

-   classification of variables <https://www.ncbi.nlm.nih.gov/pmc/articles/PMC7312700/>

-   more on the definition of well-being in PISA <https://files.eric.ed.gov/fulltext/ED612544.pdf>

## Settings

```{r}
#loaded librarires
library(dplyr)
library(psych) #for KMO test and principal()
library(car) #to apply transformations
library(MVN) #to perform multivariate gaussianity check
library(GGally) #for ggcorr
library(ggplot2)
```

```{r}
#DIRECTORIES
root_proj_dir = "../../"
dataset_dir = paste(root_proj_dir,"/data/pisa_wPV_grouped_bysch.csv",sep="")
include_dir = paste(root_proj_dir,"/src/include/Utilities.R",sep="")
#INCLUDING UTILITIES
source(include_dir)
#IMPORTING THE DATASET
pisa_data <- read.csv(file=dataset_dir)
```

```{r}
#some adjustments on the data
pisa_data$X <- NULL
pisa_data$schID <- NULL
pisa_data$CNT <- as.factor(pisa_data$CNT)
pisa_data$CNTSCHID <- as.factor(pisa_data$CNTSCHID)
pisa_data$CNTSTUID <- as.factor(pisa_data$CNTSTUID)
```

# Data preprocessing

```{r}
#paluesible values dataset
plausible_values <- pisa_data %>% select(starts_with("PV"))
#excluding PV frmo the dataset to compute the factor analysis
pisa_data <- pisa_data %>% select(-starts_with("PV")) #excluding target variables
```

```{r}
#standardize the variables
transformed_data <- as.data.frame(scale(select_if(pisa_data,is.numeric)))
transformed_data$CNT <- pisa_data$CNT #adding CNT column

pisa_data <- transformed_data
rm(transformed_data)

boxplot(select_if(pisa_data,is.numeric),las=2)
```

# Well-being variables

```{r}
grouped_variables <- list()

#Psychological well-being: c("EUDMO","SWBP","GFOFAIL", "RESILIENCE", "COMPETE")
grouped_variables[["Psychological Well-being"]] <- c("EUDMO","SWBP", "RESILIENCE")

#Social well-being: c("EMOSUPS","BELONG","BEINGBULLIED","PERCOOP","PERCOMP")
grouped_variables[["Social Well-Being"]] <- c("EMOSUPS","BELONG","BEINGBULLIED")

#all together
grouped_variables[["Overall Well-Being"]] <- c("EUDMO","SWBP", "RESILIENCE", "COMPETE","EMOSUPS","BELONG","BEINGBULLIED","PERCOOP")

#groups
group_list <- names(grouped_variables)
```

```{r}
p_ <- GGally::print_if_interactive

for(group_name in group_list){
  
  p_(ggcorr(pisa_data[,grouped_variables[[group_name]]],
         label=TRUE,label_size = 2) +
    ggtitle(group_name))
}
```

# Estimating the model

## KMO ratios

```{r}
for(group in names(grouped_variables)){
  cat("\nGroup: ", group,"\n\n")
  print(KMO(pisa_data[,grouped_variables[[group]]]))
}
```

## Selecting number of factors

```{r}
nfactors <- list()

for(group in group_list){
  #parallel analysis
  parallel.an <- fa.parallel(pisa_data[,grouped_variables[[group]]], 
              main = paste("Parallel Analysis Scree Plots:",group))
  
  nfactors[[group]]$FA <- parallel.an$nfact
  nfactors[[group]]$PC <- parallel.an$ncomp
}
```

Fixes

```{r}
#Psychological well-being
nfactors$`Psychological Well-being`$FA <- 1
nfactors$`Psychological Well-being`$PC <- 1

#Social well-being
nfactors$`Social Well-Being`$FA <- 1
nfactors$`Social Well-Being`$PC <- 1

#Overall well-being
nfactors$`Overall Well-Being`$FA <- 2
nfactors$`Overall Well-Being`$PC <- 2
```

## Fitting

```{r}
# FA with principal factor method
fit_FA_MinRes <- list() 
# rotated PCA
fit_PCA_rotated <- list()

#fitting the models
for(group in group_list){
  
  #FA with MinRes
  fit_FA_MinRes[[group]] <- fa(r = cor(pisa_data[,grouped_variables[[group]]]),
                            nfactors = nfactors[[group]]$FA,
                            rotate = "oblimin")
  #advanced PCA
  fit_PCA_rotated[[group]] <- principal(r = cor(pisa_data[,grouped_variables[[group]]]),
                               nfactors = nfactors[[group]]$PC, 
                               rotate = "oblimin")
}
```

# Interpretation

## Loadings

```{r}
for(group in group_list){
  
  #EFA
  for(i in 1:nfactors[[group]]$FA){
    barplot(fit_FA_MinRes[[group]]$loadings[,i], 
            main = paste(group,"- EFA: Factor",i),las=2,cex.names=0.7)
  }
  #rotated PCA
  for(i in 1:nfactors[[group]]$PC){
    barplot(fit_PCA_rotated[[group]]$loadings[,i], 
            main = paste(group,"- PCA: Rotated Component",i),las=2,cex.names=0.7)
  }
}
```

Thresholdings the loadings to compute alpha

```{r}
threshold <- 0.35

significant_vars <- list()

for(group in group_list) {
  #EFA
  significant_vars[[group]]$FA <- list()
  for(i in 1:nfactors[[group]]$FA) {
    selected_indeces <- ifelse(abs(fit_FA_MinRes[[group]]$loadings[,i]) < threshold, FALSE, TRUE)
    significant_vars[[group]]$FA[[i]] <- grouped_variables[[group]][selected_indeces]
  }
  #rotated PCA
  significant_vars[[group]]$PC <- list()
  for(i in 1:nfactors[[group]]$PC) {
    selected_indeces <- ifelse(abs(fit_PCA_rotated[[group]]$loadings[,i]) < threshold, FALSE, TRUE)
    significant_vars[[group]]$PC[[i]] <- grouped_variables[[group]][selected_indeces]
  }
}
```

## GOF

```{r}
for(group in group_list){
  cat("\n-----------Group:", group, "-----------\n")
  
  #alpha score
  #cat("\nalpha:\n")
  #alpha <- psych::alpha(pisa_data[,grouped_variables[[group]]], check.keys = T)
  #cat(alpha$total[1]$raw_alpha)
  
  #number of factors and components
  cat("\n\nNumber of factors: ", nfactors[[group]]$FA)
  cat("\nNumber of components: ", nfactors[[group]]$PC)
  
  #MinRes factor analysis
  cat("\n\n### MinRes Factor Analysis ###\n")
  
  #communalities
  cat("-> Communalities:\n")
  print(fit_FA_MinRes[[group]]$communality)
  
  #fit score
  cat("-> Fit of the model\n", 
      factor.fit(r = cor(select_if(pisa_data[,grouped_variables[[group]]],is.numeric)),
             fit_FA_MinRes[[group]]), "\n")
  
  #variance explained
  cat("-> Variance explained:\n")
  print(fit_FA_MinRes[[group]]$Vaccounted)
  
  #factors internal consistency
  cat("-> Factor internal consistency:\n")
  
  for(i in 1:nfactors[[group]]$FA){
    alpha <- psych::alpha(pisa_data[,significant_vars[[group]]$FA[[i]]], 
                          check.keys = T)$total[1]
    cat("\nFactor ", i, "\n")
    print(alpha$raw_alpha)
  }
  
  #rotated PCA
  cat("\n\n### Rotated PCA ###\n")
  
  #communalities
  cat("-> Communalities:\n")
  print(fit_PCA_rotated[[group]]$communality)
  
  #fit score
  cat("-> Fit of the model: ", fit_PCA_rotated[[group]]$fit,"\n")
  cat("-> Variance explained:\n")
  print(fit_PCA_rotated[[group]]$Vaccounted)
  
  #Component internal consistency
  cat("-> Component internal consistency:\n")
  
  for(i in 1:nfactors[[group]]$PC){
    alpha <- psych::alpha(pisa_data[,significant_vars[[group]]$PC[[i]]], 
                          check.keys = T)$total[1]
    cat("\nComponent ", i, "\n")
    print(alpha$raw_alpha)
  }
}
```

### Alpha score

```{r}
#Psychological well-being c("EUDMO","SWBP","GFOFAIL", "RESILIENCE", "COMPETE")
#Social well-being c("EMOSUPS","BELONG","BEINGBULLIED","PERCOOP","PERCOMP")

#Psychological Well-Being
psych::alpha(pisa_data[, c("EUDMO","SWBP","RESILIENCE","COMPETE")], check.keys = T)$total
psych::alpha(pisa_data[, c("EUDMO","SWBP","RESILIENCE","COMPETE")], check.keys = T)$alpha.drop

#Social Well-Being
psych::alpha(pisa_data[, c("EMOSUPS","BELONG","BEINGBULLIED","PERCOOP")], 
             check.keys = T)$total
psych::alpha(pisa_data[, c("EMOSUPS","BELONG","BEINGBULLIED","PERCOOP")], 
             check.keys = T)$alpha.drop

#Overall Well-being
psych::alpha(pisa_data[, c("EUDMO","SWBP", "RESILIENCE", "COMPETE","EMOSUPS","BELONG","BEINGBULLIED","PERCOOP")], 
             check.keys = T)$total
psych::alpha(pisa_data[, c("EUDMO","SWBP", "RESILIENCE", "COMPETE","EMOSUPS","BELONG","BEINGBULLIED","PERCOOP")], 
             check.keys = T)$alpha.drop
```

# Computation of the indeces

```{r eval=FALSE, include=FALSE}
SocialWB <- fit_PCA_rotated[["Social Well-Being"]]$loadings
PsychologicalWB <- fit_PCA_rotated[["Psychological Well-being"]]$loadings

save(SocialWB, file = "SocialWB_loadings.RData")
save(PsychologicalWB, file = "PsychologicalWB_loadings.RData")
```
